---
title: "Data"
format:
  html:
    toc: true
    code-fold: true
---

# Data

In this section, we provide an overview of the dataset used in our heart disease prediction task, along with key preprocessing and exploratory analysis steps.

# Dataset Description

We use the **Heart Disease Dataset**, which contains `n = 1025` observations and `p = 14` features, including:

- `age`: Age in years  
- `sex`: Biological sex (1 = male, 0 = female)  
- `cp`: Chest pain type (0 to 3, categorical) 
  - 0 = Typical angina  
  - 1 = Atypical angina  
  - 2 = Non-anginal pain  
  - 3 = Asymptomatic
- `trestbps`: Resting blood pressure  
- `chol`: Serum cholesterol (mg/dl)  
- `fbs`: Fasting blood sugar > 120 mg/dl (1 = true, 0 = false)  
- `restecg`: Resting ECG results  
- `thalach`: Maximum heart rate achieved  
- `exang`: Exercise-induced angina (1 = yes, 0 = no)  
- `oldpeak`: ST depression induced by exercise  
- `slope`, `ca`, `thal`: Additional heart-related features  
- `target`: Presence of heart disease (1 = yes, 0 = no)
---

# Preprocessing

We perform the following preprocessing steps:

- Convert categorical variables to factors (`sex`, `cp`, `fbs`, `restecg`, `exang`, `slope`, `ca`, `thal`)
- Rename `target` to `HeartDisease` for clarity
- Check for missing values
- Split into 70% training / 30% testing using `initial_split()` from **rsample**

---

## Data Loading and Preprocessing

```{r}
library(tidyverse)
# Load the dataset
heart <- read_csv("data/heart.csv")

# Rename variables for clarity
heart <- heart %>%
  rename(
    Age = age,
    Sex = sex,
    ChestPainType = cp,
    RestingBP = trestbps,
    Cholesterol = chol,
    FastingBS = fbs,
    RestingECG = restecg,
    MaxHR = thalach,
    ExerciseAngina = exang,
    Oldpeak = oldpeak,
    Slope = slope,
    NumMajorVessels = ca,
    Thalassemia = thal,
    HeartDisease = target
  ) %>%
  mutate(
    Sex = as.factor(Sex),
    ChestPainType = as.factor(ChestPainType),
    RestingECG = as.factor(RestingECG),
    ExerciseAngina = as.factor(ExerciseAngina),
    Thalassemia = as.factor(Thalassemia),
    HeartDisease = as.factor(HeartDisease)
  )

```
# Exploratory Data Analysis

## Age Distribution by Heart Disease Status
```{r}
# Basic structure
glimpse(heart)

# Proportion of heart disease cases
heart %>%
  count(HeartDisease) %>%
  mutate(prop = n / sum(n))

```
```{r}
# Distribution of Age
ggplot(heart, aes(x = Age, fill = HeartDisease)) +
  geom_histogram(bins = 30, position = "dodge") +
  labs(title = "Age Distribution by Heart Disease Status")

```

```{r}
# Chest Pain Type vs Disease
ggplot(heart, aes(x = ChestPainType, fill = HeartDisease)) +
  geom_bar(position = "fill") +
  labs(title = "Chest Pain Type and Heart Disease", y = "Proportion")

```

As shown, patients with heart disease (label 1) tend to be older.  
The highest heart disease rate occurs around ages 55 to 65.

Likewise, chest pain types 1 and 2 are more common in patients with heart disease, suggesting these features could be predictive.

These patterns will guide our model-building decisions and feature interpretation strategies in the next sections.

## Data Structure

We first inspect the structure of the dataset:

```{r}
glimpse(heart)
```

## Missing Value Check

There are no missing values in this dataset:

```{r}
colSums(is.na(heart))
```

## Data Loading and Preparation Code

The dataset was loaded and renamed as follows:

```{r}
heart <- read_csv("data/heart.csv") %>%
  rename(
    Age = age,
    Sex = sex,
    ChestPainType = cp,
    RestingBP = trestbps,
    Cholesterol = chol,
    FastingBS = fbs,
    RestingECG = restecg,
    MaxHR = thalach,
    ExerciseAngina = exang,
    Oldpeak = oldpeak,
    Slope = slope,
    NumMajorVessels = ca,
    Thalassemia = thal,
    HeartDisease = target
  ) %>%
  mutate(
    Sex = as.factor(Sex),
    ChestPainType = as.factor(ChestPainType),
    RestingECG = as.factor(RestingECG),
    ExerciseAngina = as.factor(ExerciseAngina),
    Thalassemia = as.factor(Thalassemia),
    HeartDisease = as.factor(HeartDisease)
  )
```

## Outcome Variable Distribution

```{r}
heart %>%
  ggplot(aes(x = HeartDisease)) +
  geom_bar(fill = "tomato") +
  labs(title = "Heart Disease Diagnosis", x = "Heart Disease (1 = Yes)", y = "Count")
```